\documentclass[11pt]{article}

\usepackage{amsmath,amssymb}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{mathtools}

\newcommand{\US}{\mathrm{US}}
\newcommand{\UK}{\mathrm{UK}}
\newcommand{\logit}{\operatorname{logit}}
\newcommand{\ind}[1]{\mathbf{1}\!\left\{#1\right\}}
\newcommand{\E}{\mathbb{E}}

\title{Relative Predictors and Logistic Model for Future Power}
\date{}

\begin{document}
\maketitle
\vspace{-1.5em}

\section{Construct relative predictors (US vs.\ UK)}

Let $t$ index years. For any variable $X$, define the \emph{log ratio}
\[
x_t \equiv \ln\!\frac{X_t^{\US}}{X_t^{\UK}},
\]
and for shares in $(0,1)$ use \emph{log-odds}.

\paragraph{Affordability (wheat).}
Given $A_t^{\US}=p_t^{\US}/y^{pc,\US}_t$ and $A_t^{\UK}$,
\[
a_t \equiv \ln\!\frac{A_t^{\US}}{A_t^{\UK}}
\quad\text{(higher $\,\Rightarrow\,$ wheat less affordable in US vs UK).}
\]
Often flip the sign so ``cheaper in US'' is positive: $\tilde a_t \equiv -a_t$.

\paragraph{GDP (level, not per capita).}
\[
g_t \equiv \ln\!\frac{\mathrm{GDP}^{\US}_t}{\mathrm{GDP}^{\UK}_t}.
\]

\paragraph{Military/industrial capacity (relative global shares).}
Let $S^{\US}_t$ and $S^{\UK}_t$ denote the US and UK shares of global military capacity:
\[
S^{\US}_t = \frac{\mathrm{MilEx}^{\US}_t}{\sum_{i} \mathrm{MilEx}^i_t}, \quad
S^{\UK}_t = \frac{\mathrm{MilEx}^{\UK}_t}{\sum_{i} \mathrm{MilEx}^i_t},
\]
where $\mathrm{MilEx}^i_t$ is military expenditure for country $i$ at time $t$. Then:
\[
m_t \equiv \ln\!\frac{S^{\US}_t}{S^{\UK}_t}
\quad\text{(log ratio of global military shares).}
\]
This captures relative position in the global military hierarchy rather than simple bilateral comparison. Alternative: use military personnel shares $\mathrm{MilPer}^i_t$ instead of expenditures.

\paragraph{Education (e.g., avg.\ years, enrollment, literacy).}
\[
e_t \equiv \ln\!\frac{\mathrm{Edu}^{\US}_t}{\mathrm{Edu}^{\UK}_t}.
\]

\paragraph{Urbanization share.}
For $s^{\US}_t, s^{\UK}_t\in(0,1)$ use a log-odds differential:
\[
u_t \equiv \underbrace{\ln\!\frac{s^{\US}_t}{1-s^{\US}_t}}_{\logit(s^{\US}_t)}
-\underbrace{\ln\!\frac{s^{\UK}_t}{1-s^{\UK}_t}}_{\logit(s^{\UK}_t)}.
\]
(A plain log ratio $\ln(s^{\US}_t/s^{\UK}_t)$ also works, but log-odds behaves better near $0/1$.)

\medskip
\noindent Add lags to reduce simultaneity: use $x_{t-\ell}$ with $\ell\in\{1,2\}$.

\section{Define a binary ``future power'' target}

Pick a horizon $h$ (e.g., 5 or 10 years). Let the \emph{relative power score} be $R_t$:
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{GDP leader (PTT-lite):} \quad $R_t \equiv \ln\!\big(\mathrm{GDP}^{\US}_t/\mathrm{GDP}^{\UK}_t\big)$.
  \item \textbf{Composite (optional):} \quad $R_t = w_g\,g_t + w_m\,m_t + w_u\,u_t$ (standardize components, set $w$ by PCA or equal weights).
\end{itemize}

Two binary targets:
\begin{align*}
\textbf{(A) Leader at } t+h:\quad
&Y^{(h)}_t = \ind{R_{t+h}\ge 0} 
\quad(=1 \text{ if US leads at } t+h).\\
\textbf{(B) Overtake in } (t,t+h]:\quad
&Y^{(h)}_t = \ind{R_t<0 \ \land\ R_{t+h}\ge 0}.
\end{align*}
(A) predicts who leads at horizon $h$; (B) predicts overtake risk within the window. With two countries and $\sim$110 years, (A) is simpler and more stable.

\section{Logistic specification}

Let $X_t = \big[\tilde a_{t-\ell},\, g_{t-\ell},\, m_{t-\ell},\, e_{t-\ell},\, u_{t-\ell}\big]$. Include a smooth time trend $f(t)$ (e.g., cubic spline) to absorb global drift:
\[
\Pr\!\big(Y^{(h)}_t=1 \mid X_t\big)
= \Lambda\!\Big(\beta_0 + \beta_a\,\tilde a_{t-\ell} + \beta_g\,g_{t-\ell}
+ \beta_m\,m_{t-\ell} + \beta_e\,e_{t-\ell} + \beta_u\,u_{t-\ell} + f(t)\Big),
\]
where $\Lambda(z)=\frac{1}{1+e^{-z}}$.

\paragraph{Interpretation.}
$\beta_g>0$: a higher US/UK GDP ratio today raises the probability the US leads at $t+h$.
$\beta_a>0$: cheaper wheat in the US (relative affordability positive) raises that probability (structural-change channel).
Expect $\beta_m>0$, $\beta_u>0$, $\beta_e>0$.

\paragraph{Trend.}
$f(t)$: natural cubic spline with 3--5 df. With only two countries, do not include country fixed effects (collinear).

\paragraph{Estimation.}
No weights needed; use robust SEs (HC1).

\section{Output: a predictive variable of future relative power}

The model returns
\[
\hat q^{(h)}_t \equiv \widehat{\Pr}\!\big(Y^{(h)}_t=1\mid X_t\big)\in[0,1],
\]
interpretable as the probability the US is the leader at $t+h$ (or the overtake risk in the hazard version).

If a smooth \emph{power score} is preferred, map the same index via a rescaled logistic:
\[
S^{(h)}_t \equiv 2\,\hat q^{(h)}_t - 1 \in [-1,1],
\]
so $>0$ favors US, $<0$ favors UK.

\section{Practical fitting details (tiny panel)}

\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Horizon/lag grid:} try $h\in\{5,10\}$, $\ell\in\{1,2\}$.
  \item \textbf{Regularization:} with $\sim$100 obs, add ridge penalty:
  \[
  \max_{\beta}\ \sum_t \ell\big(\beta;Y_t,X_t\big) - \lambda\|\beta\|_2^2.
  \]
  \item \textbf{Validation:} time-based CV (rolling origin) to avoid look-ahead.
  \item \textbf{Robustness:} swap $m_t$ across military expenditure shares (MilEx) and military personnel shares (MilPer); test composite measures combining both.
\end{itemize}

\section{Minimal data pipeline (pseudocode)}

\begin{enumerate}[leftmargin=1.5em]
\item \textbf{Build relative predictors:}
\[
\tilde a_t = -\ln\!\frac{A^{\US}_t}{A^{\UK}_t},\quad
g_t = \ln\!\frac{\mathrm{GDP}^{\US}_t}{\mathrm{GDP}^{\UK}_t},\quad
m_t = \ln\!\frac{S^{\US}_t}{S^{\UK}_t},
\]
where $S^{\US}_t = \mathrm{MilEx}^{\US}_t/\sum_i \mathrm{MilEx}^i_t$ and $S^{\UK}_t = \mathrm{MilEx}^{\UK}_t/\sum_i \mathrm{MilEx}^i_t$,
\[
e_t = \ln\!\frac{\mathrm{Edu}^{\US}_t}{\mathrm{Edu}^{\UK}_t},\quad
u_t = \logit(s^{\US}_t) - \logit(s^{\UK}_t).
\]

\item \textbf{Create target $Y^{(h)}_t$:}
\[
R_{t+h} = g_{t+h}\quad\text{(or composite)},\qquad
Y^{(h)}_t = \ind{R_{t+h}\ge 0},
\]
then drop the last $h$ years.

\item \textbf{Assemble design} with lag $\ell$, add spline $f(t)$.

\item \textbf{Fit} logistic (optionally ridge), \textbf{predict} $\hat q^{(h)}_t$.

\item \textbf{Use} $\hat q^{(h)}_t$ as the future power predictor; date ``inflections'' where $\hat q^{(h)}_t$ crosses $0.5$ or where $\Delta \hat q^{(h)}_t$ peaks.
\end{enumerate}

\end{document}
